Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2019-12-18 article/Old versions/JAS/PCADSC_article_JAS_resub_jan2020.tex
old changes to meta files

\documentclass[a4paper,12pt]{article}

\usepackage{geometry}
\geometry{margin=1in}

\usepackage[english]{babel}
\usepackage{authblk}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{mathrsfs}
%\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\usepackage{apalike}
%\usepackage[utf8x]{inputenc}
\usepackage[utf8]{inputenc}
%\usepackage{nameref,hyperref}
% line numbers
%\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\usepackage{multirow}
\usepackage[table]{xcolor}
\definecolor{lightgray}{HTML}{e6e6e6}

@@

\usepackage{setspace}
\doublespacing

@@

\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\renewcommand{\baselinestretch}{2}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2018}

@@

%\runninghead{PCADSC}

@@

\author[1]{}
%\author[1]{Anne Helby Petersen}
%\author[1]{Bo Markussen}
%\author[3]{Karl Bang Christensen}
%\affil[1]{Department of Public Health, University of Copenhagen, Copenhagen, Denmark}
%\affil[2]{Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
\date{}
\maketitle

%\corrauth{Anne Helby Petersen}
%\email{ahpe@sund.ku.dk}

\renewcommand{\abstractname}{Abstract}
\renewcommand{\figurename}{Figure}
\renewcommand{\tablename}{Table}

@@

%% Include all macros below

\section*{Introduction}\label{sec:introduction}

Data comparability is a classical topic in statistics and a recurring one in psychometrics. Often, psychometric data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically. This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration, or the order of survey questionnaire items are applied. While a mixture of modes of administration, for example mail and telephone, can improve response rates in surveys, it can also be quite problematic by inducing differences in response behavior that may lead to biased results \cite{Brambilla1987,McHorney1994}. Powers, Mishra, and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle. Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu (2016), who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions).\nocite{Liu2016} But mixing several data collection methods is an essential part of the psychometric methodology, as it allows us to simultaneously build on existing methods and answer new, empirical questions. However, if we do not address the question of comparability among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

@@

We propose three visual tools for data structure comparisons, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The use of PCA implies that we can only compare datasets with numerical variables and that we focus on the linear aspect of the  data structures. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}. Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is ignored e.g. when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.

\section*{PCA-based tools for data structure comparisons}\label{sec:pcadscintro}

PCADSC compares the covariance matrices of two subsets of a dataset. If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets. We refer to Appendix S1 for a brief introduction to PCA.

@@

\subsection*{Using PCA for data structure comparisons}
%PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and eigenvalues are meaningful and interesting objects to compare across such different datasets.

The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
$$\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}$$
 be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
\item Form the principal component analyses
\begin{align*}
S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
\end{align*}
thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
\begin{equation*}
\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
\end{equation*}
This identity will simplify some expressions below. We note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

@@

In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix S2. In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

@@

\subsection*{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
\end{align*}
Due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are. Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does. By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more influential components drive the visual impression.

@@

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

@@

\subsection*{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align*}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align*}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$. 
If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation. Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.

@@

\subsection*{A data structure comparison workflow}
%\begin{figure}
%\center
%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=5cm, align=center,
%	minimum height=2cm,
%  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
%  sq node/.style={rectangle,draw,font=\small\normalfont}]
%  \node[sq node] (1) {Input: Two datasets $\mathbf{X} \in \RR^{n_x \times d}$ \\ and $\mathbf{Y} \in \RR^{n_y \times d}$};
%  \node[sq node, below = 2 cm of 1] (2) {Make CE plot};
%  \node[sq node, below left of = 2] (3) {Make angle plot};
%  \node[sq node, below right of = 2] (4) {Stop and conclude that \\ the datasets are \textit{not} \\ similar in structure};
%  \node[sq node, below left of = 3] (5) {Stop and conclude that  \\ the datasets \textit{are} \\ similar in structure};
%  \node[sq node, below right of = 3] (6) {Make chroma plot: \\  Explain dissimilarities};
%  \path[every node/.style={sloped, anchor = south, font=\small}]
%    (1) edge node  {} (2)
%    (2) edge node {Similarity} (3)
%    (2) edge node {Dissimilarity} (4)
%    (3) edge node {Similarity} (5)
%    (3) edge node {Dissimilarity} (6);
% \end{tikzpicture}
% \caption{The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper %understanding of similarities and dissimilarities in the data structures.}
% \label{fig:flowchart}
% \end{figure}

@@

\section*{Comparing psychological well-being: A data example}\label{sec:dataexample}

>>>>

% interacttfssample.tex
% v1.05 - August 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

%\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled
%\setlength\bibindent{2em}% To increase hanging indent in bibliography when line spacing is doubled

\usepackage[numbers,sort&compress]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{[}{]}{,}{n}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

\usepackage{microtype}
\usepackage{multirow}
\usepackage[table]{xcolor}
\definecolor{lightgray}{HTML}{e6e6e6}

%\usepackage{caption}
%\captionsetup{font={stretch=2}}

@@

%\usepackage{setspace}
%\doublespacing

@@

\makeatletter
\newcounter{savesection}
\newcounter{apdxsection}
\renewcommand\appendix{\par
  \setcounter{savesection}{\value{section}}%
  \setcounter{section}{\value{apdxsection}}%
  \setcounter{subsection}{0}%
  \gdef\thesection{\@Alph\c@section}}
\newcommand\unappendix{\par
  \setcounter{apdxsection}{\value{section}}%
  \setcounter{section}{\value{savesection}}%
  \setcounter{subsection}{0}%
  \gdef\thesection{\@arabic\c@section}}
\makeatother

%\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
%\renewcommand{\baselinestretch}{2}

@@

\articletype{Application Note}% Specify the article type or omit as appropriate

@@

%\author{
%\name{Anne Helby Petersen\textsuperscript{a}\thanks{CONTACT Anne Helby Petersen. Email: ahpe@sund.ku.dk}, Bo Markussen\textsuperscript{b} and Karl Bang Christensen\textsuperscript{a}}
%\affil{\textsuperscript{a}Department of Public Health, University of Copenhagen, Copenhagen, Denmark; \textsuperscript{b}Department of Mathematical Sciences, University of Copenhagen, Copenhagen, Denmark}
%}

\maketitle

@@

\begin{keywords}
Principal component analysis, exploratory data analysis, covariance matrix,
data structure 
\end{keywords}
\title{}

\section{Introduction}\label{sec:introduction}

Data comparability is a recurring topic in a wide array of statistical applications. Often, data will be collected in such a way that it is essentially divided into several subsets whose comparability needs to be assessed empirically. This happens for instance when data are collected across several centers (or countries) or when different versions of an instrument, the mode of administration of a survey, or the order of survey questionnaire items are applied. While a mixture of modes of administration, for example mail and telephone, can improve response rates in surveys, it can also be quite problematic by inducing differences in response behavior that may lead to biased results \cite{Brambilla1987,McHorney1994}. Powers, Mishra, and Young report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful \cite{Powers2005}, so the comparability issue is not just a statistical puzzle. Similarly, combining data obtained from different sampling schemes can also be problematic, as illustrated by Liu, who warns against combining online panel data with intercept samples (a pool of respondents obtained through banners, ads, or promotions)\cite{Liu2016}. But mixing several data collection methods is essential for acheiving further advances in surveying- and data collection methodology, as it allows us to simultaneously build on existing methods and answer new empirical questions. However, if we do not address the question of comparability among the data subsets arriving from e.g. different countries, we risk conducting analyses whose most fundamental assumptions are not satisfied.

@@

We propose three visual tools for data structure comparisons, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods use principal component decomposition of the empirical covariance matrix in the two subsets to create intuitive visualizations of data structure differences. The use of PCA implies that we can only compare datasets with numerical variables and that we focus on the linear aspect of the  data structures. The proposed tools are implemented in the \texttt{R} package \texttt{PCADSC} \cite{PCADSC}. %Below, we describe the procedures and present a worked data example using open source, online available data on psychological well-being in three European countries from the European Social Survey. More specifically, we compare data from Denmark with data from Bulgaria and Sweden, respectively, to investigate whether or not data on psychological well-being can be combined across countries - a question that is ignored e.g. when international rankings of happiness among countries are posted by the UN \textit{World Happiness Report} project \cite{WHR2016}.

\subsection{Comparing psychological well-being}
An example of where the data structure comparability issue must be considered is when conducting international rankings of countries according to e.g. educational quality (e.g. the PISA project) and citizen happiness (e.g. the UN \textit{World Happiness Report} project). From a methodological point of view, such international rankings are problematic, as they rely on the fundamental assumption that the measured concepts are inherently the same across countries.

We use data from the 2012 version of the European Social Survey (ESS) project to investigate inter-country differences in psychological well-being and happiness.
In the rankings of happiness, not much work has yet been devoted to evaluating the assumption of international comparability, though Veenhoven presents a theoretically thorough, but empirically simplistic, summary of possible reasons for lack of comparability \cite{Veenhoven2012}, while Lolle \& Andersen show highly potent translation issues for the term \textit{happiness} \cite{Lolle2016}. If for instance two countries differ in terms of how social networks are typically built and structured, with one emphasizing family relations and the other mostly focusing on other social relations, then having a weak family connection does not have the same implications in the first country as it does in the second one. More specifically, whereas in the first country, lack of familial network might be related to loneliness, lack of general social capital, and isolation, in the second country, the quality of the family network might not be informative at all about other aspects of a person's social or psychological well-being. The two countries thus differ in how different aspects or measures of psychological well-being are interrelated, which is essentially a difference in data structures. And therefore, comparing the two countries in these measures is not a meaningful endeavour.

Our starting point is Denmark, a small, Northern European country that has repeatedly been awarded with the title of ``happiest country in the world''  by the \textit{World Happiness Report}, most recently in 2016 \cite{WHR2016}, and we wish to investigate if this title is really meaningful at all. In order to do this, we compare the Danish ESS psychological well-being data with that of Bulgaria. Though both countries are European and thus neither very far apart geographically nor culturally, these two countries have previously been highlighted to be very different in terms of what defines happiness \cite{ESStopline5}. Moreover, intra-European, regional differences in the relationship between social capital and happiness have also been demonstrated \cite{Rodriguez2014}. When comparing Northern European countries to other European countries, a much less pronounced relationship between the two concepts is found. In particular, interpersonal relations play a less important role in Denmark, compared to Bulgaria. Therefore, a successful method for data comparisons should be able to detect these differences by looking at data on psychological well-being from these two countries. We also compare the Danish data with Swedish data in order to investigate if the PCADSC tools actually do hold discriminatory power. Denmark and Sweden are both Northern European countries and are often deemed very similar in terms of culture and history. Therefore, we expect fundamental concepts, such as psychological well-being, to be similar across these two countries.

\section{Methodology}\label{sec:pcadscintro}

PCADSC compares the covariance matrices of two subsets of a dataset. If all variables in the subsets are jointly normal with known means, the covariance matrices are sufficient statistics describing the joint distribution of all the variables, but even without the normality assumption, pairwise correlations and marginal variable variances are still interesting quantities describing linear interrelations between variables. This makes the empirical covariance matrix a reasonable place to start looking for differences in data structures. Computing empirical covariance matrices for the two subsets and comparing them entry-by-entry becomes increasingly difficult as the number of variables increases and for this reason we decompose and recompose the covariance matrices using \textit{principal component analysis} (PCA) to get an overview of differences between the two subsets. %We refer to Appendix \ref{appendix.introPCA} for a brief introduction to PCA.

@@

\subsection{Using PCA for data structure comparisons}
%PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and eigenvalues are meaningful and interesting objects to compare across such different datasets.

The tools presented here are all based on comparing the PCA results across two different datasets that contain the same variables, $\mathbf{X}$ and $\mathbf{Y}$. Let 
\begin{equation*}
\mathbf{Z} = \begin{pmatrix} \mathbf{X} \\ \mathbf{Y} \end{pmatrix} \in \RR^{(n_x + n_y) \times d}
\end{equation*}
 be the combined dataset. For each of these three datasets, we complete the following steps (here described for $\mathbf{X}$ only):
\begin{enumerate}
\item Standardize each of the variables to have mean zero and unit standard deviation. Let $\tilde{\mathbf{X}}_{j} \in \RR^{n_x \times d}$ be the standardized dataset.
\item Form the principal component analyses
\begin{align*}
S_x &= \frac{1}{n_x-1} \sum_{i=1}^{n_x} \tilde{x}_{i \cdot} \tilde{x}_{i \cdot}^\top = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top 
\end{align*}
thereby obtaining loadings $\eta_{j}^x$ and eigenvalues $\lambda^x_j$ for $j=1,\dotsc,d$.
\end{enumerate}
The hereby obtained PCA decompositions of the correlation matrices can then be compared. The standardization implies that the diagonal elements of $S_x$, $S_y$, and $S_z$ all equal 1, and thus also that
\begin{equation*}
\sum_{j=1}^d \lambda_j^x = \sum_{j=1}^d \lambda_{j}^y = \sum_{j=1}^d \lambda_{j}^z =  d.
\end{equation*}
This identity will simplify some expressions below. We note that the sequence of loadings $\eta_j$ and their associated eigenvalues $\lambda_j$ yield a simultaneous description of the structure of the dataset for all approximating dimensions $q$. This implies that the loadings and the eigenvalues can be used to investigate the structure of the dataset without the need to decide on an approximating dimension, $q$, a priori.

@@

In \textit{model 1} the observed variables $V_3$ and $V_4$ are caused by a common, unobserved variable $L_1$. Further details on the models and the simulated data are available in the supplementary material, including covariance matrices for the two models, in Appendix \ref{appendix.simData}. In dataset $\mathbf{A}$, all observations stem from the same, underlying data structure and thus, the PCADSC plots should point towards no notable differences in data structures. In dataset $\mathbf{B}$, on the other hand, there are two fundamentally different, underlying data structures and therefore, the PCADSC plots should illustrate this lack of homogeneity in data structures.

@@

\subsection{The cumulative eigenvalue plot}
The cumulative eigenvalue (CE) plot compares the eigenvalues of the correlation matrices. These eigenvalues represent how much information is withheld in each component in terms of explained variance. Thus, by comparing cumulative sums of eigenvalues, it is possible to obtain a detailed picture of how the two datasets differ in terms of what components are the most informative. In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
&(0,0), \;
(\lambda_1^z,\lambda_{1}^x-\lambda_{1}^y), \;
(\lambda_1^z + \lambda_2^z,\lambda_{1}^x+\lambda_{2}^x-\lambda_{1}^y-\lambda_{2}^y), \;
\ldots, \\
&\bigg( \sum_{j=1}^d \lambda_j^z, \sum_{j=1}^d \lambda_{j}^x - \sum_{j=1}^d \lambda_{j}^y \bigg)
\end{align*}
Due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the  first axis, the less alike the cumulative eigenvalue sums for the two datasets are. Moreover, a positive cumulative difference implies that dataset $\mathbf{X}$ holds more information in the first components than dataset $\mathbf{Y}$ does. By using the cumulative eigenvalues of $\mathbf{Z}$ as the first coordinate, we obtain a visualization where a component take up an amount of horizontal space that corresponds to the variance explained by the component (that is, its eigenvalue), thereby letting the more influential components drive the visual impression.

@@

\subsection{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

@@

\subsection{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align*}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align*}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$. 
If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation. Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.

@@

\subsection{A data structure comparison workflow}
%\begin{figure}
%\center
%\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto, node distance=5cm, align=center,
%	minimum height=2cm,
%  thick,main node/.style={rectangle, rounded corners,draw,font=\small\normalfont},
%  sq node/.style={rectangle,draw,font=\small\normalfont}]
%  \node[sq node] (1) {Input: Two datasets $\mathbf{X} \in \RR^{n_x \times d}$ \\ and $\mathbf{Y} \in \RR^{n_y \times d}$};
%  \node[sq node, below = 2 cm of 1] (2) {Make CE plot};
%  \node[sq node, below left of = 2] (3) {Make angle plot};
%  \node[sq node, below right of = 2] (4) {Stop and conclude that \\ the datasets are \textit{not} \\ similar in structure};
%  \node[sq node, below left of = 3] (5) {Stop and conclude that  \\ the datasets \textit{are} \\ similar in structure};
%  \node[sq node, below right of = 3] (6) {Make chroma plot: \\  Explain dissimilarities};
%  \path[every node/.style={sloped, anchor = south, font=\small}]
%    (1) edge node  {} (2)
%    (2) edge node {Similarity} (3)
%    (2) edge node {Dissimilarity} (4)
%    (3) edge node {Similarity} (5)
%    (3) edge node {Dissimilarity} (6);
% \end{tikzpicture}
% \caption{The PCADSC workflow, starting with two datasets and then successively applying the three visual tools for obtaining a deeper and deeper %understanding of similarities and dissimilarities in the data structures.}
% \label{fig:flowchart}
% \end{figure}

@@

\section{Application}\label{sec:dataexample}
